Metagenomic Barcoding of Pollen Loads Offers Insights on the Foraging Patterns of Queen Bumble Bees

### 2023-02-20 script/manuscript_MEE.Rmd
more cleaning

--- 
title: Metagenomic Barcoding of Pollen Loads Offers Insights on the Foraging Patterns of Queen Bumble Bees
author:  |
    | Reed Clark Benkendorf$^1$$^,$$^2$^[Correspondence: reedbenkendorf2021@u.northwestern.edu], Jane E. Ogilive$^3$$^,$$^4$, Emily J. Woodworth$^1$$^,$$^2$,
    |  Sophie Taddeo$^1$$^,$$^2$, Paul J. CaraDonna$^1$$^,$$^2$$^,$$^3$, Jeremie B. Fant$^1$$^,$$^2$  
    |  $^1$Chicago Botanic Garden, 1000 Lake Cook Road, Glencoe, Illinois 60022, USA  
    |  $^2$Plant Biology and Conservation, Northwestern University, Evanston, Illinois 60208, USA  
    |  $^3$ Rocky Mountain Biological Laboratory, P.O. Box 519, Crested Butte, Colorado 81224, USA
    |  $^4$ Australia post fire project, ----, ----, ----, AU  
abstract:  |  
  |  
  1) DNA Barcoding has been successful for the rapid analysis of ecological assemblages. Although barcoding in the plant kingdom barcoding has been more difficult than others, and hence may begin to lag behind other kingdoms. 
  2) Here we test the utilization of Angiosperms 353 probes to barcode plant species found in pollen loads collected from Queen Bumble Bees. 
  3) To verify the accuracy for this barcoding system we compared the data to museum species, observation studies, and species distribution modelling to identify likely candidate species. 
  4) By utilizing Species distribution modelling we allow users to create a regionally appropriate sequence databases which may use increase the alignment algorithms minimizing need for large computational power, and run time. 
  5) We show that the Angiosperms 353 probes, which are currently being used in the largest ever plant systematic endeavor, offers significant promise to metagenomic approaches. 
  6) Understanding plants in ecological contexts and understandings of their synecology.
keywords: |
  DNA baroding, metagenomics, hyb-seq, Angiosperms353, plant metagenomics, bombus queens, subalpine, species distribution model
output:
  pdf_document: default
  toc: no
  word_document: default
csl: "../citations/methods-in-ecology-and-evolution.csl"
bibliography: ../citations/citations.bib
link-citations: yes
fig_caption: yes
always_allow_html: yes
header-includes:
- \usepackage{endfloat}
- \usepackage{setspace}\doublespacing
- \usepackage{lineno}
- \linenumbers
---

@@

rm(corb, week_tab, flr_date_ranges, pollen_sample_info, blst)

@@

    sample.id == 3 & Genus == 'Delphinium' ~ plant.species,
    sample.id == 4 & Genus %in% ranunc_splash ~ 'Delphinium.nuttallianum',
    sample.id == 4 ~ plant.species, # 4 has a viola split

    sample.id == 6 ~ plant.species, # has a delphinium split

    sample.id == 9 ~ plant.species, # has both Delphinium
    sample.id == 10 ~ plant.species,

    sample.id == 14 ~ plant.species,

    sample.id == 18 & Genus == 'Cynoglossum' ~ 'Mertensia.fusiformis',
    sample.id == 18 ~ plant.species,

    sample.id == 19 & Genus == 'Nemophila' ~ 'Hydrophyllum.fendleri', 
    sample.id == 19 ~ plant.species,
    sample.id == 20 & TAXON_NEW == 'Hydrophyllum canadense' ~ 'Hydrophyllum fendleri',
    sample.id == 20 ~ plant.species,
    sample.id == 21 ~ plant.species,

    sample.id == 23 & Genus %in% ranunc_splash ~ 'Delphinium.nuttallianum',
    sample.id == 23 ~ plant.species, # has both Viola
    sample.id == 24 & Genus %in% ranunc_splash ~ 'Delphinium.nuttallianum',
    sample.id == 24 ~ plant.species,
    sample.id == 25 & Genus %in% ranunc_splash ~ 'Delphinium.nuttallianum',
    sample.id == 25 ~ plant.species,
    sample.id == 26 & Genus %in% ranunc_splash ~ 'Delphinium.nuttallianum',
    sample.id == 26 ~ plant.species,
    sample.id == 27 & Genus %in% ranunc_splash ~ 'Delphinium.nuttallianum',
    sample.id == 27 ~ plant.species, # has both Hydrophyllum
    sample.id == 28 & Genus %in% ranunc_splash ~ 'Delphinium.nuttallianum',
    sample.id == 28 ~ plant.species,
    sample.id == 29 ~ plant.species,
    sample.id == 30 & Genus == 'Cynoglossum' ~ 'Mertensia.fusiformis', 
    sample.id == 30 ~ plant.species,
    sample.id == 32 & Genus %in% ranunc_splash ~ 'Delphinium.nuttallianum',
    sample.id == 32 ~ plant.species,
    sample.id == 33 ~ plant.species, # has both Viola

    sample.id == 35 & Genus %in% ranunc_splash ~ 'Delphinium.nuttallianum',
    sample.id == 36 & Genus == 'Cynoglossum' ~ 'Mertensia.fusiformis',  # 7.99
    sample.id == 35 ~ plant.species, # has both Viola
    sample.id == 36 ~ plant.species,

    sample.id == 39 & Genus %in% ranunc_splash ~ 'Delphinium.nuttallianum',
    sample.id == 39 ~ plant.species, # has both hydrophyllum
    sample.id == 40 ~ plant.species,
    sample.id == 41 & Genus == 'Cynoglossum' ~ 'Mertensia.ciliata', 
    sample.id == 41 ~ plant.species,
    sample.id == 42 ~ plant.species,
    sample.id == 43 ~ plant.species,
    sample.id == 44 ~ plant.species, # 44 has both delphinium

    sample.id == 46 & Genus == 'Cynoglossum' ~ 'Mertensia.fusiformis', 
    sample.id == 46 ~ plant.species,

    sample.id == 49 & Genus %in% ranunc_splash ~ 'Delphinium.barbeyi',
    sample.id == 49 ~ plant.species,

    sample.id == 51 & Genus %in% ranunc_splash ~ 'Delphinium.nuttallianum',
    sample.id == 51 ~ plant.species, # 51 has a viola split
    sample.id == 52 ~ plant.species,
    sample.id == 54 ~ plant.species,

    sample.id == 55 & Genus == 'Cynoglossum' ~ 'Mertensia.fusiformis', 
    sample.id == 55 ~ plant.species,

    sample.id == 56 & Genus == 'Nemophila' ~ 'Hydrophyllum fendleri',
    sample.id == 56 ~ plant.species,

    sample.id == 60 ~ plant.species,
    sample.id == 65 ~ plant.species, # 65 has a mertensia split

   TRUE ~ as.character(TAXON_NEW)
  )) %>% 
  mutate(
    TAXON_EXPERT = str_replace_all(TAXON_EXPERT, '[.]', ' '),
    TAXON_EXPERT = if_else(is.na(TAXON_EXPERT), TAXON_NEW, TAXON_EXPERT),
    TAXON_EXPERT = str_replace(TAXON_EXPERT, 'Geum ternatum', 'Geum triflorum'),
    TAXON_EXPERT = str_replace(
      TAXON_EXPERT, 'Arenaria globiflora|Minuartia recurva|Sagina procumbens', 'Eremogone congesta'),
    TAXON_EXPERT = if_else(doy < 171 & Genus == 'Cynoglossum', 
                           'Mertensia fusiformis', TAXON_EXPERT), 
    TAXON_EXPERT = if_else(doy < 190 & TAXON_EXPERT == 'Lupinus argenteus', 
                           'Lupinus sericeus', TAXON_EXPERT)) %>%

@@

# re class the totals to sum to 100%

@@

@@

  Tandem to the lack of continued expertise required to generate and maintain species lists, is the expertise required to continue tracking when major phenological events occur in many plant species at relatively fine scales or under novel climates. Knowledge of these events is currently limited to general time periods of only a handful of phenological events and groups of organisms (e.g. flowering initiation, or trees) (@prather2004implications, @li2016responses). While many programs and initiatives exist to collect phenological information on subsets of easily identifiable charismatic species to detect major trends in phenology, these capture only a subset of the extent diversity (@betancourt2005implementing, @havens2007chicago). In many instances it appears that while landscapes respond similarly to environmental variables which predict phenological responses, that individual species vary widely in their responses to similar environmental cues, or respond to different cues (@augspurger2020concordance, @xie2015deciduous, @xie2018predicting, @caradonna2014shifts).  **As can be seen here, predictions of when a single, major phenological event occurs is already data limited, with sample size having an effect on the subset of species which we could even generate weibull estimates for. ; check assumptions of model again and just do a quasi something** A more promising approach for the tropics may lay in circular statistics (@park2022herbarium).

>>>>

--- 
title: Metagenomic Barcoding of Pollen Loads Offers Insights on the Foraging Patterns of Queen Bumble Bees
author:  |
    |  Sophie Taddeo$^1$$^,$$^2$, Paul J. CaraDonna$^1$$^,$$^2$$^,$$^3$, Jeremie B. Fant$^1$$^,$$^2$  
    |  $^1$Chicago Botanic Garden, 1000 Lake Cook Road, Glencoe, Illinois 60022, USA  
    |  $^2$Plant Biology and Conservation, Northwestern University, Evanston, Illinois 60208, USA  
    |  $^3$ Rocky Mountain Biological Laboratory, P.O. Box 519, Crested Butte, Colorado 81224, USA
    |  $^4$ Australia post fire project, ----, ----, ----, AU  
    | Reed Clark Benkendorf$^1$$^,$$^2$^[Correspondence: reedbenkendorf2021@u.northwestern.edu], Jane E. Ogilive$^3$$^,$$^4$, Emily J. Woodworth$^1$$^,$$^2$,
abstract:  |  
  |  
  1) DNA Barcoding has been successful for the rapid analysis of ecological assemblages. Although barcoding in the plant kingdom barcoding has been more difficult than others, and hence may begin to lag behind other kingdoms. 
  2) Here we test the utilization of Angiosperms 353 probes to barcode plant species found in pollen loads collected from Queen Bumble Bees. 
  3) To verify the accuracy for this barcoding system we compared the data to museum species, observation studies, and species distribution modelling to identify likely candidate species. 
  4) By utilizing Species distribution modelling we allow users to create a regionally appropriate sequence databases which may use increase the alignment algorithms minimizing need for large computational power, and run time. 
  5) We show that the Angiosperms 353 probes, which are currently being used in the largest ever plant systematic endeavor, offers significant promise to metagenomic approaches. 
  6) Understanding plants in ecological contexts and understandings of their synecology.
keywords: |
  DNA baroding, metagenomics, hyb-seq, Angiosperms353, plant metagenomics, bombus queens, subalpine, species distribution model
output:
  word_document: default
  toc: no
  pdf_document: default
csl: "../citations/methods-in-ecology-and-evolution.csl"
bibliography: ../citations/citations.bib
link-citations: yes
fig_caption: yes
always_allow_html: yes
header-includes:
- \usepackage{endfloat}
- \usepackage{setspace}\doublespacing
- \usepackage{lineno}
- \linenumbers
---

@@

rm(corb, week_tab, flr_date_ranges, pollen_sample_info)

@@

    sample.id == 3 & Genus == 'Delphinium' ~ plant.species,
    sample.id == 4 & Genus %in% ranunc_splash ~ 'Delphinium.nuttallianum',
    sample.id == 4 ~ plant.species,                               ## COMPLETE ##

    sample.id == 6 ~ plant.species, # has a delphinium split
    # Symphyotrichum foliaceum check on day 173

    sample.id == 9 ~ plant.species, # has both Delphinium # Symphyotrichum foliaceum check on day 176

    sample.id == 10 ~ plant.species, # has many late season taxa

    sample.id == 14 ~ plant.species,  # # Symphyotrichum foliaceum check on day 165

    sample.id == 15 ~ plant.species, ## has many mysterious Rosids @ Gothic Spires  - call potentilla?

    sample.id == 18 & Genus == 'Cynoglossum' ~ 'Mertensia.fusiformis', # & Ericaceae
    sample.id == 18 ~ plant.species, # Erigeron grandiflorus on day 166 @ Frogs Pond
    sample.id == 19 & Genus == 'Nemophila' ~ 'Hydrophyllum.fendleri', # multiple Onagraceae
    sample.id == 19 ~ plant.species, # 
    sample.id == 20 & TAXON_NEW == 'Hydrophyllum canadense' ~ 'Hydrophyllum fendleri',
    sample.id == 20 ~ plant.species, # Erigeron grandiflorus doy 165 @ Avery Wash
    sample.id == 21 ~ plant.species, # Erigeron grandiflorus & cirsium parryi doy 165 @ Gothic Spires

    sample.id == 23 & Genus %in% ranunc_splash ~ 'Delphinium.nuttallianum',
    sample.id == 23 ~ plant.species, # has both Viola             ## COMPLETE ##
    sample.id == 24 & Genus %in% ranunc_splash ~ 'Delphinium.nuttallianum',
    sample.id == 24 ~ plant.species,                              ## COMPLETE ##
    sample.id == 25 & Genus %in% ranunc_splash ~ 'Delphinium.nuttallianum',
    sample.id == 25 ~ plant.species,                              ## COMPELTE ##
    sample.id == 26 & Genus %in% ranunc_splash ~ 'Delphinium.nuttallianum',
    sample.id == 26 ~ plant.species,                              ## COMPLETE ##
    sample.id == 27 & Genus %in% ranunc_splash ~ 'Delphinium.nuttallianum',
    sample.id == 27 ~ plant.species, # has both Hydrophyllum      ## COMPLETE ##
    sample.id == 28 & Genus %in% ranunc_splash ~ 'Delphinium.nuttallianum',
    sample.id == 28 ~ plant.species,                              ## COMPLETE ##
    sample.id == 29 ~ plant.species,                              ## COMPLETE ##
    sample.id == 30 & Genus == 'Cynoglossum' ~ 'Mertensia.fusiformis', 
    sample.id == 30 ~ plant.species,                              ## COMPLETE ##
    sample.id == 30 & Genus == 'Erigeron' ~ 'Erigeron elatior',
    sample.id == 32 & Genus %in% ranunc_splash ~ 'Delphinium.nuttallianum',
    sample.id == 32 ~ plant.species,                              ## COMPLETE ##
    sample.id == 33 ~ plant.species, # has both Viola
    # Symphyotrichum foliaceum check on day 166

    sample.id == 35 & Genus %in% ranunc_splash ~ 'Delphinium.nuttallianum',
    sample.id == 35 ~ plant.species, # has both Viola             ## COMPLETE ##

    sample.id == 36 & Genus == 'Cynoglossum' ~ 'Mertensia.fusiformis',  
    sample.id == 36 ~ plant.species,                              ## COMPLETE ##

    sample.id == 39 & Genus %in% ranunc_splash ~ 'Delphinium.nuttallianum',
    sample.id == 39 ~ plant.species, # has both hydrophyllum      ## COMPLETE ##
    sample.id == 40 ~ plant.species,                              ## COMPLETE ##
    sample.id == 41 & Genus == 'Cynoglossum' ~ 'Mertensia.ciliata', 
    sample.id == 41 & Genus == 'Viola' ~ 'Viola adunca',          ## COMPLETE ##
    sample.id == 41 ~ plant.species,                              ## COMPLETE ##
    sample.id == 42 ~ plant.species,
    sample.id == 43 ~ plant.species,                             ## COMPLETE  ##
    sample.id == 44 ~ plant.species, # 44 has both delphinium

    sample.id == 45 ~ plant.species,         ## EARLY ASTER DAY 163 @ FROGS.POND ##
    sample.id == 46 & Genus == 'Cynoglossum' ~ 'Mertensia.fusiformis', 
    sample.id == 46 ~ plant.species,         ## EARLY ASTER DAY 163 @ FROGS.POND ##

    sample.id == 49 & Genus %in% ranunc_splash ~ 'Delphinium.barbeyi',
    sample.id == 49 & Genus == 'Viola' ~ 'Viola.adunca',
    sample.id == 49 ~ plant.species,                              ## COMPLETE ##

    sample.id == 51 & Genus %in% ranunc_splash ~ 'Delphinium.nuttallianum',
    sample.id == 51 ~ plant.species,  ## Helianthella quinquenervis @ DOY 166 ##
    sample.id == 52 ~ plant.species,    ## Erigeron elatior  on DOY 163 @ AVERY WASH ##
    sample.id == 54 ~ plant.species,                             ## COMPLETE ##

    sample.id == 55 & Genus == 'Cynoglossum' ~ 'Mertensia.fusiformis', 
    sample.id == 55 ~ plant.species,                              ## COMPLETE ##

    sample.id == 56 & Genus == 'Nemophila' ~ 'Hydrophyllum fendleri',
    sample.id == 56 ~ plant.species,          ## DOY 169 FROGS.POND Symphyotrichum foliaceum
                                              ## & Erigeron elatior ##
    sample.id == 60 & Genus == 'Viola' ~ 'Viola.adunca',          ## COMPLETE ##
    sample.id == 60 ~ plant.species,
    sample.id == 60 & Genus == 'Viola' ~ 'Viola',                 ## COMPLETE ##
    sample.id == 65 ~ plant.species,

   TRUE ~ as.character(TAXON_NEW)
  )) %>% 
  mutate(
    TAXON_EXPERT = str_replace_all(TAXON_EXPERT, '[.]', ' '),
    TAXON_EXPERT = if_else(is.na(TAXON_EXPERT), TAXON_NEW, TAXON_EXPERT),
    TAXON_EXPERT = str_replace(TAXON_EXPERT, 'Holodiscus argenteus', 'Potentilla pulcherrima'),
    TAXON_EXPERT = str_replace(TAXON_EXPERT, 'Scabrethia scabra', 'Wyethia arizonica'),
    TAXON_EXPERT = str_replace(TAXON_EXPERT, 'Geum ternatum', 'Geum triflorum'),
    TAXON_EXPERT = str_replace(TAXON_EXPERT, 'Erigeron grandiflorus', 'Erigeron elatior'),
    TAXON_EXPERT = str_replace(TAXON_EXPERT, 'Agastache pallidiflora', 'Pedicularis'),
    TAXON_EXPERT = str_replace(TAXON_EXPERT, 'Lithophragma parviflorum', 'Lithophragma glabrum'),
    TAXON_EXPERT = str_replace(TAXON_EXPERT, 
                               'Arenaria globiflora|Minuartia recurva|Sagina procumbens', 'Eremogone congesta'),
    TAXON_EXPERT = if_else(doy <= 171 & Genus == 'Cynoglossum', 
                           'Mertensia fusiformis', TAXON_EXPERT), 
    TAXON_EXPERT = if_else(doy < 190 & TAXON_EXPERT == 'Lupinus argenteus', 
                           'Lupinus sericeus', TAXON_EXPERT))  %>%

@@

# re class the totals percent of all sequences to sum to 100%

@@

yargh <- split(blst_reclass_expert, f = blst_reclass_expert$sample.id)

@@

  Tandem to the lack of continued expertise required to generate and maintain species lists, is the expertise required to continue tracking when major phenological events occur in many plant species at relatively fine scales or under novel climates. Knowledge of these events is currently limited to general time periods of only a handful of phenological events and groups of organisms (e.g. flowering initiation, or trees) (@prather2004implications, @li2016responses). While many programs and initiatives exist to collect phenological information on subsets of easily identifiable charismatic species to detect major trends in phenology, these capture only a subset of the extent diversity (@betancourt2005implementing, @havens2007chicago). In many instances it appears that while landscapes respond similarly to environmental variables which predict phenological responses, that individual species vary widely in their responses to similar environmental cues, or respond to different cues (@augspurger2020concordance, @xie2015deciduous, @xie2018predicting, @caradonna2014shifts). As can be seen here, predictions of when a single, major phenological event occurs is already data limited. A more promising approach for the tropics may lay in utilizing circular statistics (@park2022herbarium).
